\pkg{Rgbp}: An R Package for Hierarchical Modeling and Method Checking

### 2013-06-26 jss/article.tex
changes

%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an R package for estimating and validating a two-level model (a random-effects mixed model). The estimation procedure utilizes Bayesian machinery and the validation involves checking frequency properties of the procedure via repeated sampling (which we call ``method checking"). It is found that \pkg{Rgb} characterized by such Bayesian-frequentist reconciliation has strong points of both sides, working well in small samples and yielding good coverage probability. Following three examples for Gaussian, Poisson and Binomial data support this argument.}
\Keywords{hierarchical model, multilevel model, random effects mixed model, method checking, coverage probability, Normal, Poisson, Binomial,  shrinkage, \proglang{R}}
\Plainkeywords{keywords, comma-separated, not capitalized, r} %% without formatting
%% at least one keyword must be supplied

@@

\section[Multilevel Structure]{Multilevel Structure} \label{multilevel}
A two-level or multilevel model, also called a conditionally independent hierarchical model \cite{1989}, is a very powerful tool for exploring the hierarchical structure in data. For example, we can imagine that there exists a district-level hierarchy (bigger population) for observed school-level data, or a state-level hierarchy for observed hospital data in a certain state.

\code{gbp}, one of the functions in \pkg{Rgbp}, fits such a hierarchical model whose first-level hierarchy has a distribution of observed data and second-level (bigger population hierarchy) has a conjugate prior distribution on the first-level parameter. The \code{gbp} function allows users to choose one of three types of multilevel models, such as Normal-Normal, Poisson-Gamma, and Binomial-Beta, based on their data sets.

\subsection[Normal-Normal]{Normal-Normal}
The following is a general Normal-Normal hierarchical model. For reference,  $V_{j}~(\equiv \sigma^{2}/n_{j})$ below is assumed to be known or to be accurately estimated, and subscript \emph{j} indicates the \emph{j}-th group among \emph{k} groups in the data set.
\begin{equation}
y_{j}\vert \mu_{j} \stackrel{ind}{\sim}\textrm{Normal}(\mu_{j}, V_{j}),
\end{equation}
\begin{equation}
\mu_{j}\vert \beta, A\stackrel{ind}{\sim}\textrm{Normal}(\mu_{0j}, A),
\end{equation}

@@

\subsection[Poisson-Gamma]{Poisson-Gamma}
\code{gbp} is also able to build a Poisson-Gamma multilevel model. Note that a constant, $1/r$, multiplied to the Gamma distribution below is a scale and a square bracket below indicates [mean, variance] of distribution. And for notational consistency, let's define $y_{j}\equiv z_{j} / n_{j}$ for all $j$.
\begin{equation}
z_{j}\vert \lambda_{j} \stackrel{ind}{\sim}\textrm{Poisson}(n_{j}\lambda_{j}),
\end{equation}
\begin{equation}
\lambda_{j}\vert \beta, r\stackrel{ind}{\sim}\frac{1}{r}\textrm{Gamma}(\lambda_{0j}r)\sim \textrm{Gamma} \left[\lambda_{0j}, ~\frac{\lambda_{0j}}{r} \right],
\end{equation}

where $\log(\lambda_{0j}) =x_{j}'\beta$, $j=1, \ldots, k$, with two unknown hyper-parameters, $r$ and $\beta$. The posterior distribution of this Poisson-Gamma model is
\begin{equation} \label{gammapost}
\lambda_{j}\vert \textbf{z}, \beta, r \stackrel{ind}{\sim}\frac{1}{r + n_{j}}\textrm{Gamma}(r\lambda_{0j} + n_{j}y_{j})\sim\textrm{Gamma} \left[\lambda^{\ast}_{j},~\frac{\lambda^{\ast}_{j}}{r+n_{j}} \right],
\end{equation}

@@

where $\log(\frac{p_{0j}}{1-p_{0j}}) =x_{j}'\beta, ~j=1, \ldots, k$ and $r$ and $\beta$ are two unknown hyper-parameters. Then, corresponding posterior distribution is
\begin{equation} \label{betapost}
p_{j}\vert \textbf{z}, \beta, r \stackrel{ind}{\sim}\textrm{Beta}(rp_{0j}+n_{j}y_{j},~r(1-p_{0j})+n_{j}(1-y_{j}))\sim\textrm{Beta}\bigg[p^{\ast}_{j},~ \frac{p^{\ast}_{j}(1-p^{\ast}_{j})}{r+n_{j}+1}\bigg],
\end{equation}
where $p^{\ast}_{j}\equiv(1-B_{j})y_{j}+B_{j}p_{0j}$, $B_{j}\equiv r/ (r+n_{j})$, and $y_{j}\equiv z_{j} / n_{j}$, $j=1,\ldots,k$.

@@

\subsection[ADM]{Adjustment for Density Maximization}
It is noted that the shrinkage factors ($B_1, \ldots B_k$) are a function of the second-level variance component, \emph{i.e.}, $B_{j}\equiv V_{j}/(V_{j}+A)=B_{j}(A)$ for Gaussian and $B_{j}\equiv r/(r+n_{j})=B_{j}(r)$ for Poisson and Binomial models. Current popular methods of estimation via maximization (MLE or MAP) of these unknown shrinkage factors can result in estimates lying on the boundary of the parameter space. In the Normal-Normal model \cite{2012} this situation typically arises when the within-group variation is greater than the between-group variation resulting in a parameter estimate of $\hat{A} = 0$.

@@

% TODO: Here, Christiansen and Morris (1997) seems appropriate format
There are two kinds of model justification processes; one is a model checking and the other is a method checking. The model checking is for the justification of the model \ref{multilevel} on a specific data set. One possible question is, ``Can this data set benefit from such a multilevel modeling?'' \cite{1997} answered this question by using the Negative-Binomial mixture model on Poisson data to justify the second-level hierarchy. They found that their data had more variation than expected of the first-level Poisson distribution and Poisson hierarchical model could successfully account for such additional variation.

Once we are sure that the hierarchical modeling can be appropriate for our data, the following question will be about the validity of interval estimates, the final product of this multilevel modeling. ``Does the 95\% (can be specified differently) confidence interval obtained via this Bayesian model-fitting process achieve 95\% confidence level  for any true parameter values?'' \pkg{Rgbp} has a function called \code{coverage} to answer this question and it comprises of two parts, generating peudo-dataset and estimating coverage probability.

For reference, the explanation will be based on Normal-Normal model because the idea can be easily applied to other two models.

\subsection{Pseudo-data Generation Process}
Figure 1 will be helpful to understand this process. As we can see in (\ref{normalpost}), the distribution of each true parameter ($\mu_{j},~j=1,\ldots, k$) is completely determined by two hyper-parameters, $A$ and $\beta_{(m\times1)}$. So, once we fix these hyper-parameters at specific values, we can generate true parameters. Suppose we sampled 500 $\mathbf{\mu}_{(k\times1)}$'s, \emph{i.e.}, $\{\mathbf{\mu}^{(i)}_{(k\times1)},~i=1, \ldots, 500\}$ from the prior distribution in (2), where $A$ and $\beta$ are given. Then, we can also generate $\{\mathbf{y}^{(i)}_{(k\times1)},~i=1, \ldots, 500\}$ given each $\mu_{ij},$ where $i$ indicates the $i$-th pseudo-data set and $j$ does the $j$-th group.

@@

\subsection{Coverage Estimation Process}
Next, \code{coverage} fits the Normal-Normal model 500 times using the 500 peudo-datasets in order to obtain 500$\times$k interval estimates.

@@

For this purpose, you gathered data of 31 hospitals in NY composed of the number of deaths ($z$) within a month of CABG surgeries and total number of patients ($n$) receiving CABG surgeries in each hospital. The following code is an example of input based on the last ten hospital data.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c( 14,   9,  15,  13,  35,  26,  25,  20,   35,   27)
R> n <- c(593, 602, 629, 636, 729, 849, 914, 940, 1193, 1340)
\end{CodeInput}
\end{CodeChunk}

@@

Assuming an additional hierarchy is reasonable, a model-fitting process begins.  Since the true death rate per exposure after CABG surgery might be small and the caseload ($n_{j}$) is much bigger than the number of deaths ($z_{j}$) in each hospital, the Poisson distribution would be the first choice to describe the uncertainty in our data. Note that caseload  ($n_{j}$) can be interpreted as an exposure, which is not necessarily an integer.

Next, \code{gbp} will help us fit the Poisson multilevel model with the Gamma conjugate prior distribution on the true death rate $\lambda_{j}$ whose mean is 0.020 ($\lambda_{0}=0.020$) as described in 2.2 . For reference, the number of regression coefficients ($m$) is 0 because we do not need to estimate the prior mean via log-linear regression (see section 3.2).
\begin{CodeChunk}
\begin{CodeInput}
R> data(hospital)
R> z <- hospital$z
R> n <- hospital$n
R> p <- gbp(z, n, mean.PriorDist = 0.02, model = "pr")
R> p
\end{CodeInput}
\begin{CodeOutput}
Summary for each unit (sorted by n):

@@

Intuitively, the result of multilevel modeling makes more sense than that of naive sample mean. For example, suppose there are two hospitals, whose sample means ($z_{j} / n_{j}$) are 0 and 0.001 and caseloads ($n_{j}$) are 1 and 1000 each. Do you believe that the former hospital, whose observed death rate is 0, is better than the latter and are you going to choose the former hospital? Borrowing information from state-level hierarchy seems reasonable for the former hospital because it is hard to judge its true death rate per exposure with just one caseload. Though somewhat extreme, this is what happened to the two left-most hospitals in the first plot and this is why hierarchical modeling is a reasonable choice for this data set.

The estimated 95\% intervals are displayed on the right-side plot. We can clearly see that all the posterior means (red dots) are between sample mean (empty dots) and second-level mean (a blue horizontal line). For reference, if we want to draw this plot by the order of data input, \code{plot(p, sort = FALSE)} is a right adjustment. Overall, as the caseload increases towards the right and as the posterior mean gets closer to 0, the length of interval gets shorter, which the formula of posterior variance in (6) implies.

@@

Then, how reliable are these intervals? Does our procedure to generate interval estimates have good repeated sampling property? The following method checking will answer this question. The \code{coverage} function below generates 10,000 pseudo-data sets regarding the estimated $r~(=336.37)$ as a given true value. For reference, we can try any other value of $r$, for example $r=200$, by replacing below code with `\code{R> pcv <- coverage(p, A.or.r = 200, mean.PriorDist = 0.02, nsim = 10000)}.

@@

It estimated coverage probabilities by simple unbiased estimates (a blue dashed line) and Rao-Blackwellized unbiased estimates (a red line). Both lines are indistinguishable from the horizontal line at 0.95 and the estimated overall average coverage rate is 0.950 from the Rao-Blackwellized estimate. This result shows that the interval estimates from the suggested multilevel modeling on this particular data set have nice repeated sampling property.

@@

The data indicate that independent Binomial distribution is appropriate probability distribution for describing each player's number of hits among 45 at-bats conditioning on the unknown true batting average, \emph{i.e.}, $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(45, p_{j}), ~j=1, \ldots, 18$. For reference, we can also assume the Normal distribution using $y_{j}=z_{j} / n_{j}$ and $V_{j}=\bar{y}(1-\bar{y})/n$, where $\bar{y}=\sum_{j} z_{j} / \sum_{j} n_{j}$, $j=1,\ldots, 18$.

@@

Let's see shrinkage and 95\% interval plots for more intuition. It is obvious that sample means (empty dots) on the upper line shrink towards two different means, 0.310 and 0.233. For reference, the red line symbols on dots are for when two or more points have the same mean and are plotted over each other. For example, five players (from the 11th player to the 15th) have the same sample mean (0.222) and at this point on the upper line, there are short red lines toward five directions. This symbol also means that the seemingly two lines crossing over others were actually three lines for those who were outfielders (10th, 14th, and 15th hitters).

@@

One more thing to emphasize is that observed sample means are all within the interval estimates. And these interval estimates also covered the batting averages over the remaining season successfully, although the result was left out. Then, how much can we trust these interval estimates? The following method check will answer it.

@@

The meaning of putting the \code{gbp} object, \code{b}, without any additional argument in the above \code{coverage} function is to make this function regard the estimated values, such as $\hat{r}$ (=112.95) and $\hat{\beta}~(=(-1.194, ~0.389)^{T})$, as true values when it generates pseudo-data sets.

>>>>

%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an R package for estimating and validating a two-level model (a random-effects mixed model). The estimation procedure utilizes Bayesian machinery and the validation involves checking frequency properties of the procedure via repeated sampling (which we call ``method checking"). It is found that \pkg{Rgb} characterized by such Bayesian-frequentist reconciliation has strong points of both sides, working well in small samples and yielding good coverage probability. Following three examples for Gaussian, Poisson and Binomial data validate this point.}
\Keywords{hierarchical model, multilevel model, random effects mixed model, method checking, coverage probability, Normal, Poisson, Binomial,  shrinkage, \proglang{R}}
\Plainkeywords{keywords, comma-separated, not capitalized, r} %% without formatting
%% at least one keyword must be supplied

@@

\section[Multilevel Structure]{Multilevel Structure} \label{multilevel}
A two-level or multilevel model, also called a conditionally independent hierarchical model \cite{1989}, is a very powerful tool for exploring the hierarchical structure in data. For example, we can imagine that there exists a district-level hierarchy (bigger population) for observed school-level data, or a state-level hierarchy for observed hospital-level data in a certain state.

\code{gbp}, one of the functions in \pkg{Rgbp}, fits such a hierarchical model whose first-level hierarchy has a distribution of observed data and second-level (bigger population hierarchy) has a conjugate prior distribution on the first-level parameter. The \code{gbp} function allows users to choose one of three types of multilevel models, such as Normal-Normal, Poisson-Gamma, and Binomial-Beta, based on their datasets.

\subsection[Normal-Normal]{Normal-Normal}
The following is a general Normal-Normal hierarchical model. For reference,  $V_{j}~(\equiv \sigma^{2}/n_{j})$ below is assumed to be known or to be accurately estimated, and subscript \emph{j} indicates the \emph{j}-th group among \emph{k} groups in the dataset.
\begin{equation}
y_{j}\vert \mu_{j} \stackrel{ind}{\sim}\textrm{Normal}(\mu_{j}, V_{j}),
\end{equation}
\begin{equation}
\mu_{j}\vert \beta, A\stackrel{ind}{\sim}\textrm{Normal}(\mu_{0j}, A),
\end{equation}

@@

\subsection[Poisson-Gamma]{Poisson-Gamma}
\code{gbp} is also able to build a Poisson-Gamma multilevel model. Note that a constant, $1/r$, multiplied to the Gamma distribution below is a scale and a square bracket below indicates [mean, variance] of distribution. And for notational consistency, let's define $y_{j}\equiv z_{j} / n_{j}$ for all $j$.
\begin{equation}
z_{j}\vert \lambda_{j} \stackrel{ind}{\sim}\textrm{Poisson}(n_{j}\lambda_{j}),
\end{equation}
\begin{equation}
\lambda_{j}\vert \beta, r\stackrel{ind}{\sim}\frac{1}{r}\textrm{Gamma}(r\lambda_{0j})\sim \textrm{Gamma} \left[\lambda_{0j}, ~\frac{\lambda_{0j}}{r} \right],
\end{equation}

where $\log(\lambda_{0j}) =x_{j}'\beta$, $j=1, \ldots, k$, with two unknown hyper-parameters, $r$ and $\beta$. The posterior distribution of this Poisson-Gamma model is
\begin{equation} \label{gammapost}
\lambda_{j}\vert \textbf{y}, \beta, r \stackrel{ind}{\sim}\frac{1}{r + n_{j}}\textrm{Gamma}(r\lambda_{0j} + n_{j}y_{j})\sim\textrm{Gamma} \left[\lambda^{\ast}_{j},~\frac{\lambda^{\ast}_{j}}{r+n_{j}} \right],
\end{equation}

@@

where $\log(\frac{p_{0j}}{1-p_{0j}}) =x_{j}'\beta, ~j=1, \ldots, k$ and $r$ and $\beta$ are two unknown hyper-parameters. Then, corresponding posterior distribution is
\begin{equation} \label{betapost}
p_{j}\vert \textbf{y}, \beta, r \stackrel{ind}{\sim}\textrm{Beta}(rp_{0j}+n_{j}y_{j},~r(1-p_{0j})+n_{j}(1-y_{j}))\sim\textrm{Beta}\bigg[p^{\ast}_{j},~ \frac{p^{\ast}_{j}(1-p^{\ast}_{j})}{r+n_{j}+1}\bigg],
\end{equation}
where $p^{\ast}_{j}\equiv(1-B_{j})y_{j}+B_{j}p_{0j}$, $B_{j}\equiv r/ (r+n_{j})$, and $y_{j}\equiv z_{j} / n_{j}$, $j=1,\ldots,k$.

@@

\subsection[ADM]{Adjustment for Density Maximization}
It is noted that the shrinkage factors ($B_1, \ldots B_k$) are a function of the second-level variance component, \emph{i.e.}, $B_{j}\equiv V_{j}/(V_{j}+A)=B_{j}(A)$ for Gaussian and $B_{j}\equiv r/(r+n_{j})=B_{j}(r)$ for Poisson and Binomial models. Current popular methods of estimation via maximization (MLE or MAP) of these unknown shrinkage factors can result in estimates lying on the boundary of the parameter space. In the Normal-Normal model this situation typically arises when the within-group variation is greater than the between-group variation resulting in a parameter estimate of $\hat{A} = 0$ \cite{2012}.

@@

% TODO: Here, Christiansen and Morris (1997) seems appropriate format
There are two kinds of model justification processes; one is a model checking and the other is a method checking. The model checking is for the justification of the multilevel model on a specific dataset. One possible question is, ``Can this dataset benefit from such a multilevel modeling?'' \cite{1997} answered this question by using the Negative-Binomial mixture model on Poisson data to justify the second-level hierarchy. They found that their data had more variation than expected of the first-level Poisson distribution and Poisson hierarchical model could successfully account for such additional variation.

Once we make sure that the hierarchical modeling can be appropriate for our data, the following question will be about the validity of interval estimates, the final product of this multilevel modeling. ``Does the 95\% (can be specified differently) confidence interval obtained via this Bayesian model-fitting process achieve 95\% confidence level  for any true parameter values?'' \pkg{Rgbp} has a function called \code{coverage} to answer this question and it comprises of two parts, generating peudo-dataset and estimating coverage probability.

For reference, the explanation will be based on Normal-Normal model because the idea can be easily applied to the other two models.

\subsection{Pseudo-data Generation Process}
Figure 1 will be helpful to understand this process. As we can see in (\ref{normalpost}), the distribution of each true parameter ($\mu_{j},~j=1,\ldots, k$) is completely determined by two hyper-parameters, $A$ and $\beta_{(m\times1)}$. So, once we fix these hyper-parameters at specific values, we can generate true parameters. Suppose we sampled 500 $\mathbf{\mu}_{(k\times1)}$'s, \emph{i.e.}, $\{\mathbf{\mu}^{(i)}_{(k\times1)},~i=1, \ldots, 500\}$ from the prior distribution in (2), where $A$ and $\beta$ are given. Then, we can also generate $\{\mathbf{y}^{(i)}_{(k\times1)},~i=1, \ldots, 500\}$ given each $\mu_{ij},$ where $i$ indicates the $i$-th pseudo-dataset and $j$ does the $j$-th group.

@@

\subsection{Coverage Estimation Process}
Next, \code{coverage} fits the Normal-Normal model 500 times using the 500 peudo-datasets in order to obtain 500$\times k$ interval estimates.

@@

For this purpose, this person gathered data of 31 hospitals in NY composed of the number of deaths ($z$) within a month of CABG surgeries and total number of patients ($n$) receiving CABG surgeries in each hospital. The following code is an example of input based on the last ten hospital data.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c( 14,   9,  15,  13,  35,  26,  25,  20,   35,   27)
R> n <- c(593, 602, 629, 636, 729, 849, 914, 940, 1193, 1340)
\end{CodeInput}
\end{CodeChunk}

@@

Assuming an additional hierarchy is reasonable, a model-fitting process begins.  We know the true death rate per exposure after CABG surgery might be small and the caseload ($n_{j}$) is much bigger than the number of deaths ($z_{j}$) in each hospital. Then the independent Poisson distribution, \emph{i.e.},  $z_{j}\vert \lambda_{j}\stackrel{ind}{\sim} \textrm{Poisson}(n_{j}\lambda_{j})$, $j=1, \ldots, 31,$ would be the first choice to describe the uncertainty in our data. Note that caseload  ($n_{j}$) can be interpreted as an exposure, which is not necessarily an integer.

Next, \code{gbp} will help us fit the Poisson multilevel model with the Gamma conjugate prior distribution on the true death rate $\lambda_{j}$ whose mean is 0.020 ($\lambda_{0}=0.020$) as described in the section 2.2. For reference, the number of regression coefficients ($m$) is 0 because we do not need to estimate the prior mean via log-linear regression (see section 3.2).
\begin{CodeChunk}
\begin{CodeInput}
R> data(hospital)
R> z <- hospital$z
R> n <- hospital$n
R> p <- gbp(z, n, mean.PriorDist = 0.02, model = "pr")
R> p
\end{CodeInput}
\begin{CodeOutput}
Summary for each unit (sorted by n):

@@

Intuitively, the result of multilevel modeling makes more sense than that of naive sample mean. For example, suppose there are two hospitals, whose sample means ($z_{j} / n_{j}$) are 0 and 0.001 and caseloads ($n_{j}$) are 1 and 1000 each. Do you believe that the former hospital, whose observed death rate is 0, is better than the latter and are you going to choose the former hospital? Borrowing information from state-level hierarchy seems reasonable for the former hospital because it is hard to judge its true death rate per exposure with just one caseload. Though somewhat extreme, this is what happened to the two left-most hospitals in the first plot and this is why hierarchical modeling is a reasonable choice for this dataset.

The estimated 95\% intervals are displayed on the right-side plot. We can clearly see that all the posterior means (red dots) are between sample mean (empty dots) and second-level mean (a blue horizontal line). For reference, if we want to draw this plot by the order of data input, \code{plot(p, sort = FALSE)} is the right adjustment.

@@

Then, how reliable are these intervals? Does our procedure to generate interval estimates have good repeated sampling property? The following method checking will answer this question. The \code{coverage} function below generates 10,000 pseudo-datasets regarding the estimated $r~(=336.37)$ as a given true value. For reference, we can try any other value of $r$, for example $r=200$, by replacing below code with `\code{R> pcv <- coverage(p, A.or.r = 200, mean.PriorDist = 0.02, nsim = 10000)}.

@@

It estimated coverage probabilities by simple unbiased estimates (a blue dashed line) and Rao-Blackwellized unbiased estimates (a red line). Both lines are indistinguishable from the horizontal line at 0.95 and the estimated overall average coverage rate is 0.950 from the Rao-Blackwellized estimate. This result shows that the interval estimates from the suggested multilevel modeling on this particular dataset have nice repeated sampling property.

@@

The data indicate that independent Binomial distribution is appropriate probability distribution for describing each player's number of hits ($z_{j}$) among 45 at-bats conditioning on the unknown true batting average, \emph{i.e.}, $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(45, p_{j}), ~j=1, \ldots, 18$. For reference, we can also assume the Normal distribution using $y_{j}=z_{j} / n_{j}$ and $V_{j}=\bar{y}(1-\bar{y})/n$, where $\bar{y}=\sum_{j} z_{j} / \sum_{j} n_{j}$, $j=1,\ldots, 18$.

@@

Let's see shrinkage and 95\% interval plots for more intuition. It is obvious that sample means (empty dots) on the upper line shrink towards two different means, 0.310 and 0.233. For reference, the red line symbols on dots are for when two or more points have the same mean and are plotted over each other. For example, five players (from the 11th player to the 15th) have the same sample mean (0.222) and at this point on the upper line, there are short red lines toward five directions.

@@

Then, how much can we trust these interval estimates? The following method checking that regards the estimated values, $\hat{r}$ (=112.95) and $\hat{\beta}~(=(-1.194, ~0.389)^{T})$, as true values when it generates pseudo-datasets will answer it.

@@
